---
title: "Chapter 12 - Tidy Data"
author: "Mike McCann"
date: "10/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
```


### Tidy Data

## 12.1 Introduction

Tidying your data allows for you to focus on your analytic tasks. Spend the time to tidy your data so you don't have to come back and do it later...

Tidy Data paper from the journal of statistical software <- http://www.jstatsoft.org/v59/i10/paper

## 12.1.1 Prerequisites

## 12.2 Tidy Data

Rules of Tidy Data
1. Each variable must have its own column
2. Each observation must have its own row
3. Each value must have its own cell

Leads to two practical instructions
1. Put each dataset in a tibble
2. Put each variable in a column

Advantages of tidy data....
-consistent data structure makes it easier to apply tools/methods later across multiple data sets.
-variables in columns allows R to do its vectorised magic. 

## 12.2.1 Exercises

# 1. Using prose, describe how the variables and observations are organised in each of the sample tables.

```{r}

# Sample 1: Each variable is given its own column and each observation has its own row. Clean Data.

# Sample 2: Multiple observations for each country/year across two rows. Not Clean

# Sample 3: variables of count and population merged into a rate column. Single observation per row. Not Clean.

# Sample 4: Data spread between multiple tables. Also multiple observations per row.

```

# 2. Compute the rate for table2, and table4a + table4b. You will need to perform four operations:

-Extract the number of TB cases per country per year.
-Extract the matching population per country per year.
-Divide cases by population, and multiply by 10000.
-Store back in the appropriate place.
-Which representation is easiest to work with? Which is hardest? Why?

```{r}

table2cases <- table2 %>% 
  filter(type == "cases") %>% 
  rename(cases = count)


table2pop <- table2 %>%  
  filter(type == "population") %>% 
  rename(population = count)

table2comb <- tibble(
  country = table2cases$country,
  year = table2cases$year,
  cases = table2cases$cases,
  population = table2pop$population
)

table2final <- table2comb %>% 
  mutate(rate = (cases / population) * 10000)


```

# 3. Recreate the plot showing change in cases over time using table2 instead of table1. What do you need to do first?

```{r}
table2final %>% 
  ggplot(aes(year, cases)) +
  geom_line(aes(group = country), colour = 'grey50') +
  geom_point(aes(color = country))
```

## 12.3 Pivoting

-Most data in the wild isn't tidy. It is often organized to facilitate uses other than analysis.
-SOlve with pivot longer: Increase number of rows by decreasing columns

## 12.3.1 Longer

-Use when dataaset has column names that are not names... but variables.
-SOlve with pivot longer: Increase number of rows by decreasing columns.

-pivot_longer uses the arguements names_to and values_to
-ex: pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

## 12.3.2 Wider

-Use when observations are scattered across multiple rows.

-pivot_wider uses the arguments names_from and values_from
-ex: pivot_wider(names_from = type, values_from = count)

## 12.3.3 Exercises

# 1. Why are pivot_longer() and pivot_wider() not perfectly symmetrical? Carefully consider the following example:

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")

stocks

# the year column becomes a character data type during the transformation from wide to long. 

stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return", 
               names_ptypes = list(year = double()))

#I believe this is trying to coerce the tible to convert the column back to a double... but it throws an error. pivot_wider documentation suggests using names_transform instead

stocks %>%
  pivot_wider(names_from = year, values_from = return)%>%
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return",
               names_transform = list(year = as.numeric))

```

# 2. Why does this code fail? 
```{r}
table4a %>% 
  pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")

# Code forgot to add backticks ` to make the non-character column names 1999 and 2000 function

table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

```

# 3. What would happen if you widen this table? Why? How could you add a new column to uniquely identify each value?

```{r}
people <- tribble(
  ~name,             ~names,  ~values,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)

head(people)

# Brett:
people %>% 
  pivot_wider(names_from = names,
              values_from = values
  )
people %>% 
  mutate(obs_id = str_c(name, names, values)) %>% 
  pivot_wider(names_from = names,
              values_from = values
  )
----
  # Mike:
  people %>% 
  pivot_wider(names_from = "names", values_from = "values")

# Ends up showing the number of observations for each variable.

people %>%
  group_by(name, names) %>%
  mutate(obs = row_number()) %>% 
  pivot_wider(names_from = "names", values_from = "values")


```

# 4. Tidy the simple tibble below. Do you need to make it wider or longer? What are the variables?

```{r}

preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

preg %>%
  pivot_longer(c('male', 'female'), names_to = "sex", values_to = "count")

preg %>% 
  pivot_longer(c(male, female), names_to = "gender", values_to = "count")

```

## 12.4 Separating and Uniting

## 12.4.1 Separate

-separate() pulls apart one column into multiple columns.
-separate takes the column name and the names of the new columns 
-ex: separate(rate, into = c("cases", "population"))

-separate() will split values wherever it sees non-alphanumeric characters by default
-you can pass a specific character by which to separate using the sep argument
-ex: separate(rate, into = c("cases", "population"), sep = "/")
-sep can take regex...

-separate will keep the column data type the same. Can ask it to convert it based on best guess heuristics using convert = TRUE.

## 12.4.2 Unite

-Inverse of separate. 
-default is to unite with a "_" in between the united columns. use sep arguement to change this (ex "").

## 12.4.3 Exercises

# 1. What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets.

```{r}

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"))

#separate drops the "g" by defualt. extra = merge tells it to keep it.

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")


tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))

#likewise separate by default provides NA for missing values. fill = left/right moves where it takes from.

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")

```

# 2. Both unite() and separate() have a remove argument. What does it do? Why would you set it to FALSE?

```{r}

#remove removes the original input columns from the argument for either unite or separate.

```

# 3. Compare and contrast separate() and extract(). Why are there three variations of separation (by position, by separator, and with groups), but only one unite?

```{r}

#extract appears to work as if separate had remove = TRUE. IE it does not keep the original columns. Documentation suggests extract can also be used more easily with regex.

#Unite is taking multiple columns and combining them so you don't need to know 'how' to separate (by position, by separator, etc).

```

## 12.5 Missing Values

-Explicitly missing values are those marked/flagged with NA.
-implicitly missing values are those where it is not present from the dataset.

"An explicit missing value is the presence of an absence; an implicit missing value is the absence of a presence."

IF explicitly missing values are not important you can use values_drop_na = TRUE in pivot longer to get rid of them.

IF explicit values are important you can use complete(). complete() takes a set of columns and finds all unique combinations and inserting explicit NAs where appropriate

-fill() will analyze a column and take the most recent non-missing value and apply it. Used primarily in datasets that were previously used for data entry.

## Exercises

# 1. Compare and contrast the fill arguments to pivot_wider() and complete().

```{r}

# In pivot_wider values_fill specifies what values should be filled in when there is missing data
# in complete() fill provides a single value used instead of NA 

```

# 2. What does the direction argument to fill() do?

```{r}

# determines the direction it should fill from... ie if it takes from above or below.
# by default set to "down". can also take "up", "downup", or "updown"

treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)

treatment %>% 
  fill(person, .direction = "downup")

```

## 12.6 Case Study

```{r}
head(who)
```


## 12.7 Non-Tidy Data

