---
title: "Chapter 3 - Data Visualization"
author: "Mike McCann"
date: "7/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(maps)
library(mapproj)

```

----------
3.2.4 Exercises

Question 1: Run ggplot(data = mpg). What do you see?

```{r}

ggplot(data = mpg)

```


Question 2: How many rows are in mpg? How many columns?

```{r}
str(mpg) 

dim(mpg)
```

Question 3: What does the drv variable describe? Read the help for ?mpg to find out.

```{r}

```

Question 4: Make a scatterplot of hwy vs cyl.

```{r}
ggplot(mpg, aes(x = displ, y = cyl)) +
  geom_point() +
  geom_line(mapping = aes(x = displ, y = cty))

ggplot(data = mpg) + 
  geom_point(mapping = aes(displ, hwy))

```

Question 5: What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r}

ggplot(data = mpg)+
  geom_point(
    mapping = aes(x = class, y = drv)
  )

```

-----------
3.3.1 Exercies

Question 1: Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

```{r}

done

```

Question 2: Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r}
ggplot(data = mpg)+
  geom_point(
    mapping = aes(x = class, y = hwy, color = hwy)
  )
```

Question 3: What happens if you map the same variable to multiple aesthetics?

```{r}

ggplot(data = mpg)+
  geom_point(
    mapping = aes(x = displ, y = cty, color = displ, size = displ)
  )
```

Question 4: What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, fill = hwy), shape = 8, stroke = 1)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 0, stroke = 5)

```

Question 5: What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

```{r}

ggplot(data = mpg)+
  geom_point(
    mapping = aes(x = displ, y = hwy, color = displ < 5))

```

Question 6: What’s gone wrong with this code? Why are the points not blue?

ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ), alpha = .5)



```

-----------

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ cyl, nrow = 1)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)

```

-----------
3.5.1 Exercises

Question 1: What happens if you facet on a continuous variable?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ hwy, nrow = 2)

```


Question 2: What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?

ggplot(data = mpg) + 
geom_point(mapping = aes(x = drv, y = cyl))

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl)) +
  facet_grid(drv ~ cyl)
```

Question 3: What plots does the following code make? What does . do?

ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy)) +
facet_grid(drv ~ .)

ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy)) +
facet_grid(. ~ cyl)

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)

```

Question 4: Take the first faceted plot in this section:

ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy)) + 
facet_wrap(~ class, nrow = 2)

What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) 

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


Question 5: Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?

```{r}
# ncol telling it the number of columns when setting up the facets
# nrow telling it the number of rows when setting up the facets

# facet grid pairwise combinations so it needs all variables to work (cannot tell it how the columns/rows work)
```

Question 6: When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?

```{r}

```

-----------
3.6.1 Exercises

Question 1: What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

```{r}

# geom_line
# geom_boxplot
# geom_hist
# geom_area

```

Question 2: Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
geom_point() + 
geom_smooth(se = FALSE)

```{r}

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = T)

```

Question 3: What does show.legend = FALSE do? What happens if you remove it?
Why do you think I used it earlier in the chapter?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point(show.legend = FALSE) + 
  geom_smooth(se = TRUE, show.legend = FALSE)

```

Question 4: What does the se argument to geom_smooth() do?

```{r}

```

Question 5: Will these two graphs look different? Why/why not?

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
geom_point() + 
geom_smooth()

ggplot() + 
geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy), color = "blue") + 
  geom_point() +
  geom_smooth(color = "red")

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy), alpha = .5) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy), color = "red")

```

Question 6: Recreate the R code necessary to generate the following graphs.

```{r}

ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ, y=hwy))+
  geom_smooth(mapping = aes(x=displ, y=hwy), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(aes(group = drv), se = FALSE)

# Andrew 
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, group = drv)) +
  geom_point(mapping = aes(color = drv)) +
  geom_smooth(mapping = aes(color = drv), se = FALSE )

# Wolfgang
ggplot(data = mpg)+
  geom_point(mapping = aes(x=displ,y=hwy,color=drv))+
  geom_smooth(mapping = aes(x=displ, y=hwy, color=drv), se = FALSE)

# Drew
ggplot(mpg, aes(x = displ, y = hwy, col = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
  
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(col = drv)) +
  geom_smooth(se = FALSE)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(col = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(aes(fill = drv), color = "white", shape = 21, size = 2, stroke = 2)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 8, col = 'white') +
  geom_point(aes(col = drv), size = 4)

```


-----------

3.7.1 Exercises

1. What is the default geom associated with stat_summary()? How could you rewrite the previous plot to use that geom function instead of the stat function?

```{r 3.7.1 Question 1}

ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth)
    ,
    fun.min = min,
    fun.max = max,
    fun = median
  )

ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = cut, y = depth))


```

Question 2: What does geom_col() do? How is it different to geom_bar()?

```{r 3.7.1 Question 2}

# geom_col and geom_bar provide similar graphs... however geom_col requires both an x and a y variable. Meaning that you must provide a count argument for y to fill out the geom_col.

diamonds %>% 
  count(cut) %>% 
  ggplot() +
  geom_col(aes(cut, n))

ggplot(diamonds) +
  geom_bar(aes(cut))

```

Question 3: Most geoms and stats come in pairs that are almost always used in concert. Read through the documentation and make a list of all the pairs. What do they have in common?

```{r}

ggplot(diamonds) +
  geom_bar(aes(cut))

ggplot(diamonds) +
  stat_count(aes(cut))

```

Question 4: What variables does stat_smooth() compute? What parameters control its behaviour?

```{r}

# stat_smooth and geom_smooth take the same arguements but stat_smooth is used for non-standard geoms. Both provide a smooth line vice a more jagged line as shown below.

# ggplot(diamonds) +
#   geom_line(aes(carat, depth)) +
#   geom_smooth(aes(x = carat, y = depth))

ggplot(diamonds) +
  stat_smooth(aes(x = carat, y = depth), level = 0.15)

```

Question 5: In our proportion bar chart, we need to set group = 1. Why? In other words what is the problem with these two graphs?

```{r}

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))

# group = 1 tell the function that the entire total should = 1 from which we want the proportions applied to y.

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), group = 1))

ggplot(data = diamonds) + 
  geom_bar(aes(x = cut, y = ..count.. / sum(..count..), fill = color))
```

-----------

Section 3.8 - Position adjustments

-position = "identity" places objects exactly where they fall in the context of the graph.

ex: 
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
geom_bar(alpha = 1/5, position = "identity")

-position = "dodge" places overlapping objects directly beside one another for easier comparison within a category.

ex: 
ggplot(data = diamonds) + 
geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

-position = "fill" sets stacked bars (geom_bar) at the same height... allowing you to look at proportions between bars

ex:
ggplot(data = diamonds) + 
geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

-position = "jitter" adds noise to a geom_point plot to avoid overplotting. Can alternatively use geom_jitter

ex:
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")

-----------

3.8.1 Exercises

Question 1: What is the problem with this plot? How could you improve it?

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
geom_point()

```{r}
# add jitter through either position = "jitter" or geom_jitter

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
geom_point()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position = "jitter")

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()
```

Question 2: What parameters to geom_jitter() control the amount of jittering?

```{r}

# use the width or height arguments. default is 40% (0.4) .. example below

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(width = 0.75, height = 0.75)

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(width = 0.1, height = 23)

```

Question 3: Compare and contrast geom_jitter() with geom_count().

```{r}

# geom_count does a similar thing to geom_jitter in that it offers an alternative to overplotting... instead of adding noise geom_count increases the size of the dots to account for multiple points in a single spot. See comparison below

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_count()
```

Question 4: What’s the default position adjustment for geom_boxplot()? Create a visualisation of the mpg dataset that demonstrates it.

```{r}

# default is dodge2... this makes sense as a boxplot is by defenition laid out next to one another.

ggplot(data = mpg, mapping = aes(x = drv, y = hwy, fill = class)) + 
  geom_boxplot()

```

------------

Section 3.9 - Coordinate Systems

-coord_flip: switches x and y axes)
-coord_quickmap: sets aspect ratio correctly for maps.
-coord_polar: turns a bar chart into a pie chart.

-----------

3.9.1 Exercises

Question 1: Turn a stacked bar chart into a pie chart using coord_polar().
```{r}

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity)) +
  labs(x = NULL, y = NULL) +
  coord_polar()

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut)) +
  labs(x = NULL, y = NULL) +
  coord_polar()
```

Question 2: What does labs() do? Read the documentation.

```{r}

# labs() allows you to add titles... or eliminate titles using the NULL operator

```

Question 3: What’s the difference between coord_quickmap() and coord_map()?

```{r}

# coord_map projects the map onto a approximate sphere which slightly alters the map to be more accurate based on the curvature of the earth. 

nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()


ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_map()
```

Question 4: What does the plot below tell you about the relationship between city and highway mpg? Why is coord_fixed() important? What does geom_abline() do?

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
geom_point() + 
geom_abline() +
coord_fixed()

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
geom_point() + 
geom_abline() +
coord_fixed()


ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
geom_point() + 
geom_abline()

# geom_abline provides a reference line that procedes diagonally at a rate of 1:1. This allows us to see that the relationship is slightly greater than 1:1

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter() + 
  # geom_smooth(se = FALsE) +
  geom_abline() +
  coord_fixed()

```

--------

Section 3.10 - The Layered Grammar of Graphics

Template:
ggplot(data = <DATA>) + 
<GEOM_FUNCTION>(
mapping = aes(<MAPPINGS>),
stat = <STAT>, 
position = <POSITION>
) +
<COORDINATE_FUNCTION> +
<FACET_FUNCTION>