---
title: "Chapter 13 - Relational Data"
author: "Mike McCann"
date: "10/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(nycflights13)

```


### Relational Data

## 13.1 Introduction

Collectively, multiple tables of data are called relational data because the relations between these tables is important (vice the individual tables on their own).

Relations are defined by pairs of tables. 

Three verbs are important when dealing with relational data:
-Mutating Joins: Add observations to one data frame from matching observations in another
-Filtering Joins: Filter observations from one df based on matching observations in another
-Set Operations: treat observations as if they were a set

most likely to find relational data in relational database management systems (RDMS).

## 13.1.1 Prerequisites

## 13.2 NYCflights13

## 13.2.1 Exercises

# 1. Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?

```{r}

# To draw the route each plane flies you would need the coordinates (Lat and Long) of the origin and destination airports. You would need the flight data from the flights dataset and then join that with the lat/long from the airports dataset

```

# 2. I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?

```{r}

head(weather)
head(airports)

# The relationship between weather and airports can be found between the weather$origin and airports$faa 

```

# 3. weather only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with flights?

```{r}

# it would ideally contain information about weather at the destination as well. 

```

# 4. We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?

```{r}

# dataframe containing year, month, day, and holiday ex:

holiday <- tribble(
  ~year, ~month, ~day, ~holiday,
  2013, 12, 25, "Christmas"
)

head(holiday)
```

## 13.3 Keys

# The variable used to connect each pair within two tables is called a key. A key uniquely identifies an observation between two tables
-Primary Key: Uniquely identifies an observation in its own table
-Foreign Key: Uniquely identifies an observation in another table

-Surrogate key: a key you've added when no primary key is readily available. 

## 13.3.1 Exercises

# 1. Add a surrogate key to flights.

```{r}

flights %>% 
  mutate(surrogate = row_number())

```

# 2. Identify the keys in the following datasets
# Lahman::Batting,
# babynames::babynames
# nasaweather::atmos
# fueleconomy::vehicles
# ggplot2::diamonds

```{r}
# head(Lahman::Batting)
Lahman::Batting %>% 
  count(playerID, yearID, stint) %>% 
  filter(n > 1)

# head(babynames::babynames)
babynames::babynames %>%
  count(year, sex, name) %>% 
  filter(n > 1)

# head(nasaweather::atmos)
nasaweather::atmos %>% 
  count(lat, long, year, month) %>% 
  filter(n > 1)

# head(fueleconomy::vehicles) 
fueleconomy::vehicles %>% 
  count(id) %>% 
  filter(n > 1)

# head(ggplot2::diamonds) 
ggplot2::diamonds %>%
  count(carat, cut, color, clarity) %>% 
  filter(n > 1)

# no obvious primary key for the last one

```

# 3. Draw a diagram illustrating the connections between the Batting, People, and Salaries tables in the Lahman package. Draw another diagram that shows the relationship between People, Managers, AwardsManagers.

# How would you characterise the relationship between the Batting, Pitching, and Fielding tables?

```{r}

```

## 13.4 Mutating Joins

# Mutating joins combine variables from two tables. In effect they match observations based on a given key and copies variables from one table to another... creating new columns (just like a mutate).

## 13.4.1 Understanding Joins

## 13.4.2 Inner Join

-inner_join: joins two dataframes keeping only observations that have matches. Because it gets rid of non-matches it is not great for analysis. 

## 13.4.3 Outer Joins

-left_join: Keeps all observations from dataframe x. Unmatched from y are dropped.
-right_join: keeps all observations in dataframe y. Unmatched values from x are dropped
-full_join: keeps all observations in both x and y

## 13.4.4 Duplicate Keys

# What happens when you have keys that aren't unique??? Two possiblities..
- One table has duplicate keys. Usually indicates a one-to-many relationship
- Both tables have duplicate keys. Results in a cartesian product... not helpful.

## 13.4.5 Defining the Key Column

# Defualt of most joins (by = NULL) will use all variables that appear in both tables, called a natural join. This sometimes works but other times you will want to select your keys specifically.

# Select your own key using by = "x" to pick a single common variable to be the key. This requires the names to be the same across columns. Alternatively you can use by = c("a" = "b") if the names do not match

## 13.4.6 Exercises

# 1. Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:

```{r}
# head(flights)

flights %>% 
  group_by(dest) %>% 
  summarize(avg_delay = mean(arr_delay, na.rm = T)) %>% 
  inner_join(airports, by = c("dest" = "faa")) %>% 
  ggplot(aes(lon, lat, color = avg_delay, size = avg_delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()

```

# 2. Add the location of the origin and destination (i.e. the lat and lon) to flights.

```{r}

airports2 <- airports %>% 
  select(faa, lat, lon)

#use suffix from the documentation 

flights %>% 
  select(year, month, day, hour, minute, origin, dest) %>% 
  left_join(airports2, by = c("dest" = "faa")) %>% 
  left_join(airports2, by = c("origin" = "faa"), suffix = c("_dest", "_origin"))

```

# 3. Is there a relationship between the age of a plane and its delays?

```{r}

flights %>% 
  select(tailnum, year, arr_delay) %>% 
  left_join(planes, by = "tailnum") %>% 
  mutate(age = year.x - year.y) %>% 
  select(tailnum, arr_delay, age) %>% 
  group_by(age) %>% 
  summarize(avg_delay = mean(arr_delay, na.rm = T)) %>% 
  filter(!is.na(age)) %>% 
  ggplot() +
  geom_point(aes(age, avg_delay))


```

# 4. What weather conditions make it more likely to see a delay?

```{r}
flights %>% 
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>% 
  group_by(precip) %>% 
  summarize(avg_delay = mean(dep_delay, na.rm = T)) %>% 
  ggplot(aes(precip, avg_delay)) +
  geom_point()

flights %>% 
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>% 
  group_by(visib) %>% 
  summarize(avg_delay = mean(dep_delay, na.rm = T)) %>% 
  ggplot(aes(visib, avg_delay)) +
  geom_point()


```

# 5. What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.

```{r}

flights %>% 
  filter(year == 2013,
         month == 6,
         day == 13) %>% 
  group_by(dest) %>% 
  summarize(avg_delay = mean(arr_delay, na.rm = T)) %>% 
  inner_join(airports, by = c("dest" = "faa")) %>% 
  ggplot(aes(lon, lat, color = avg_delay, size = avg_delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()

# based on google looks like there was a series of large storms in the south and plains region.

```

## 13.4.7 Other Implementations

#base R merge can be used as a substitute for joins but is less intuitive.

#joins are based on / inspired by SQL. compare:
inner_join(x, y, by = "z")    ==	    SELECT * FROM x INNER JOIN y USING (z)
left_join(x, y, by = "z")	    ==      SELECT * FROM x LEFT OUTER JOIN y USING (z)
right_join(x, y, by = "z")    ==	    SELECT * FROM x RIGHT OUTER JOIN y USING (z)
full_join(x, y, by = "z")	    ==    SELECT * FROM x FULL OUTER JOIN y USING (z)

## 13.5 Filtering Joins

# Unlike mutating joins filtering joins remove rows instead of adding new rows

-semi_join: Keeps all observations in x that have a match in y
-anti_join: drops all observations in x that have a match in y

## 13.5.1 Exercises

# 1. What does it mean for a flight to have a missing tailnum? What do the tail numbers that don’t have a matching record in planes have in common? (Hint: one variable explains ~90% of the problems.)

```{r}
# No tailnum appears to be flights that didn't end up happening.

flights %>% 
  filter(is.na(tailnum))

#appears to be primarily MQ and AA flights. Unsure why. 

flights %>% 
  anti_join(planes, by = "tailnum")

```

# 2. Filter flights to only show flights with planes that have flown at least 100 flights.

```{r}

flights100 <- flights %>% 
  group_by(tailnum) %>% 
  count() %>% 
  filter(!is.na(tailnum), n >= 100) %>% 
  arrange(-n)

flights %>% 
  anti_join(flights100)

```

# 3. Combine fueleconomy::vehicles and fueleconomy::common to find only the records for the most common models.

```{r}

fueleconomy::vehicles %>% 
  semi_join(fueleconomy::common, by = c("make", "model"))

```

# 4. Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the weather data. Can you see any patterns?

```{r}

worsthours <- flights %>% 
  select(year, month, day, hour, origin, dep_delay) %>% 
  group_by(year, month, day, hour) %>% 
  summarize(avg_delay = mean(dep_delay, na.rm = T)) %>% 
  arrange(-avg_delay) %>% 
  head(48)

weather %>% 
  semi_join(worsthours, by = c("year", "month", "day", "hour")) %>% 
  count(visib, sort =)

```

# 5. What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?

```{r}

anti_join(flights, airports, by = c("dest" = "faa")) %>% 
  count(dest, sort = T)
# will show flights for which there is no match between dest and faa in the airports df. Possibly airports that have been recoded or miscoded?? 

anti_join(airports, flights, by = c("faa" = "dest"))
#will show airports for which there is no match between faa and dest in the flights df. This means that only airports that are not the destination of outbound flights from NYC airports.
```

# 6. You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.

```{r}

multicarrier <- flights %>% 
  select(carrier, tailnum) %>% 
  filter(!is.na(tailnum)) %>% 
  group_by(carrier, tailnum) %>% 
  count() %>% 
  ungroup() %>% 
  group_by(tailnum) %>% 
  count(tailnum, sort = T) %>% 
  filter(n >= 2)

flights %>% 
  select(carrier, tailnum) %>% 
  filter(!is.na(tailnum)) %>% 
  semi_join(multicarrier, by = "tailnum") %>% 
  group_by(carrier, tailnum) %>% 
  count() %>% 
  arrange(tailnum) %>% 
  left_join(airlines)

```


## 13.6 Join Problems

#Unfortunately our data in the wild is not likely to be as clean as the flights datasets. Here are some steps to consider when working on your own data
-Identify the variables that form a primary key in each table.
-check that none of your variables in your primary key are missing
-check that your foreign keys match primary keys in another table (anti_join helps with this)

## 13.7 Set Operaitons

# Not used frequently but still useful to know:
-intersect(x, y): returns only observations in both x and y
-union(x, y): returns unique observations in x and y
-setdiff(x, y): returns observations in x but not in y