---
title: "Chapter 7 - Exploratory Data Analysis"
author: "Mike McCann"
date: "8/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggplot2)
library(nycflights13)
library(ggstance)
library(lvplot)
library(ggbeeswarm)
```

#7.1 Introduction

## EDA Process:
## - Generate Questions about your data
## - Search for answers by visualising, transforming, and modeling your data
## - Use what you learn to refine your questions and/or generate new questions


-----

#7.2 Questions

## Use questions as a tool to guide your investigation. Questions focus your attention on a specific part of your dataset.

## Two types of questions that are always useful for discovery:
## - What type of variation occurs within my variables?
## - What type of covariation occurs betwen my variables?

## Key Terms:
## - Variable: is a quantity, quality, or property that you can measure
## - Value: is the state of a variable when you measure it. The value of a variable may change from measurement to measurement.
## - Observation: is a set of measurements made under similar conditions (you usually make all of the measurements in an observation at the same time and on the same object). An observation will contain several values, each associated with a different variable. (Sometimes refered to as a data point).
## - Tabular Data: is a set of values, each associated with a variable and an observation. Tabular data is tidy if each value is placed in its own “cell”, each variable in its own column, and each observation in its own row.

-----

#7.3 Variation

## Variation is the tendency of the values of a variable to change from measurement to measurement. 
 
## A variable is CATEGORICAL if it can only take one of a small set of values (EX: diamonds$cut).
## - Use bar chart.
## A variable is CONTINUOUS if it can take any of an infinite set of ordered values (EX: diamonds$clarity, numbers, date-time).
## - use histogram or freqpoly.

## The key to asking good follow-up questions will be to rely on your curiosity (What do you want to learn more about?) as well as your skepticism (How could this be misleading?).

#7.3.4 Exercises

# 1. Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.


```{r - 7.3.4 Question 1}
# str(diamonds)

# compare frequency across x, y, z to look for trends.
diamonds %>% 
  ggplot() + 
  geom_freqpoly(aes(x = x), color = "red", binwidth = .05) +
  geom_freqpoly(aes(x = y), color = "blue", binwidth = .05) +
  geom_freqpoly(aes(x = z), color = "green", binwidth = .05)+
  coord_cartesian(xlim = c(2, 9))

# It appears as if x and y may track with one another leading me to believe that these are likely the length and width. If this were true than z would be depth, which makes sense as it drops off after 4mm (you wouldnt want a super deep diamond...).

#confirm that x and y track together in each observation.
diamonds %>% 
  ggplot(aes (x=x, y=y)) +
  geom_point(alpha = .5) +
  geom_smooth(SE = F) +
  coord_cartesian(ylim = c(3,10),
                  xlim = c(3,10))
```


```{r - 7.3.4 Question 1}
summary(select(diamonds, x, y, z))

#Wolfgang's Solution
diamonds %>%
  filter(x > 0) %>%
  ggplot(aes(x = x, y = y)) + 
  geom_point(aes(color = z)) 

diamonds %>%
  ggplot(aes(x = x, y = z)) + 
  geom_point()

diamonds %>%
  ggplot(aes(x = y, y = z)) + 
  geom_point()

#Emily's Solution
ggplot(data = diamonds, mapping = aes(x = x)) + 
  geom_histogram(binwidth = 0.01) #between 3.5 and 9ish. Help says that x is length in mm
ggplot (data = diamonds, mapping = aes(x = y)) + 
  geom_histogram(binwidth = 0.01) #between 5 and 10. Help says that y is the width in mm 
ggplot(data = diamonds, mapping = aes(x = z))  + 
  geom_histogram(binwidth = 0.01) #between 3ish and 5ish. Help says that z is the depth in mm. Honestly, I don't know anything about diamonds so this didn't help me

```

# 2. Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

```{r - 7.3.4 Question 2}

#Emily Solution
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_histogram(binwidth = 0.5) + 
  coord_cartesian(ylim = c(0, 70)) + 
  scale_x_continuous(breaks = seq(0, 30000, by = 2500))

#Mike Solution
diamonds %>% 
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 100) +
  coord_cartesian(xlim = c(0,6000))

#Steve Observations
ggplot(filter(diamonds, price < 2500), aes(x = price)) +
  geom_histogram(binwidth = 10, center = 0)

# appears to be a major peak at ~$800 before a drop down to nearly 0 at ~$1500.
# There appears to be another small peak at roughly ~$1800.

```


# 3. How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

```{r - 7.3.4 Question 3}
diamonds %>% 
  filter(carat > .95 & carat < 1.05) %>% 
  ggplot() +
  geom_histogram(aes(x = carat), binwidth = .01)

diamonds %>% 
  filter(carat > .95 & carat < 1.05) %>% 
  count(carat)

#I would guess that the diamonds are being "rounded up" or cut to 1 carat... presumably there is a premium or prestige factor associated with reaching 1 carat. 
```


# 4. Compare and contrast coord_cartesian() vs xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?

```{r - 7.3.4 Quesiton 4}

#Brett Solution
diamonds %>% 
  # filter(y > 3 & y < 20) %>% 
  ggplot(aes(x = y)) +
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 12), ylim = c(0, 50)) 

diamonds %>% 
  # filter(y > 3 & y < 20) %>% 
  ggplot(aes(x = y)) +
  geom_histogram(binwidth = 0.5) +
  ylim(0, 50)


diamonds %>% 
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 100) +
  coord_cartesian(xlim = c(0,6000),
                  ylim = c(0, 1500))

diamonds %>% 
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 100) +
  xlim(0, 6000) +
  ylim(0, 1500)


# coord_cartesian maintains the data whereas xlim effectively filters out data outside of the given ranges. This may change your default binwidths. 

```

-----

# 7.4 Missing Values

# Usually a better practice to turn unexplained outliers into NA rather than removing them. GGPlot will drop them from plots but they will still remain in your dataset. 

## ifelse(test, yes, no). Alternate to casewhen().
## - test: logical vector
## - yes: if test is TRUE returns this value
## - no: if test is FALSE returns this value.

# 7.4.4 Exercises

# 1. What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference?

```{r - 7.3.1 Question 1}

# geom_histogram removes the NAs from the frequency calculation
diamonds %>%
  mutate(y = ifelse(y < 3 | y > 20, NA, y)) %>% 
  ggplot(aes(x = y)) +
  geom_histogram()

# geom_bar treats NA values as a distinct category and will include it as its own bar in the barchart.
flights %>%
  group_by(carrier) %>% 
  mutate(n = n()) %>% 
  ungroup() %>% 
  mutate(carrier = ifelse(n < 10000, NA, carrier)) %>% 
  ggplot() +
  geom_bar(aes(x = carrier)) 

#Kayla Example
expl_data1 <- data.frame(x1 = c(NA, 7, 8, 9, 3), # Numeric variable with one missing value
                         x2 = c(4, 1, NA, NA, 4), # Numeric variable with two missing values
                         x3 = c(1, 4, 2, 9, 6), # Numeric variable without any missing values
                         x4 = c("Hello", "I am not NA", NA, "I love R", NA)) # Factor variable with
                                                                             # two missing values
ggplot(expl_data1, aes(x4)) +
  geom_bar()

```

# 2.What does na.rm = TRUE do in mean() and sum()?

```{r - 7.4.1 Question 2}

#na.rm is NA remove. It removes the NA's from the calculation or function before executing. This is important because when NAs exist in functions like mean() and sum() it will result in NA.

#without rm.na - mean
flights %>% 
  group_by(origin) %>% 
  summarise(mean_delay = mean(dep_delay))

#with rm.na - mean
flights %>% 
  group_by(origin) %>% 
  summarise(mean_delay = mean(dep_delay, na.rm = T))

#without rm.na - sum
flights %>% 
  group_by(origin) %>% 
  summarise(mean_delay = sum(dep_delay))

#with rm.na - sum
flights %>% 
  group_by(origin) %>% 
  summarise(mean_delay = sum(dep_delay, na.rm = T))

```

-----

# 7.5 Covariation

## COVARIATION: describes behavior between variables. Tendency for the values of two or more variables to vary together in a related way.

-----

# 7.5.1 categorical and a Continuous Variable

## density adjusts geom_freqpoly to standardize count under the polygon.

## ex:
```{r - density example}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

```

## Boxplots are also good for displaying the distrobution between continuous variables broken down by a categorical variable.

# 7.5.1.1 Exercises

# 1. Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.

```{r}

#Brett SOlution
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(x = sched_dep_time, y = ..density..)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)

#Kayla Solution
flights <- flights %>% 
  mutate(cancelled = (is.na(dep_time)))

ggplot(data = flights, mapping = aes(x = sched_dep_time)) +
  geom_freqpoly(mapping = aes(color = cancelled, y = ..density..))

#Steve Solution
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot() + 
    geom_boxplot(mapping = aes(y = sched_dep_time, x = cancelled))

#Mike Solution
flights %>% 
  mutate(cancelled =  (is.na(dep_delay) | is.na(arr_delay))) %>% 
  ggplot() +
  geom_boxplot(aes(x = dep_time, y = cancelled))

```

# 2. What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?



```{r}
#Kayla Solution
ggplot(diamonds, aes(carat, price)) +
  geom_smooth(se=FALSE)

ggplot(diamonds, aes(cut, price)) +
  geom_boxplot()

ggplot(diamonds, aes(carat, price)) +
  geom_smooth(mapping = aes(color = cut)) +
  geom_point(mapping = aes(color = cut))

#Wolfgang solution
# library(GGally)
diamonds <- diamonds %>%
  mutate(diamonds, as.numeric(price))

ggally_cor(diamonds, mapping = aes(x = carat, y = price))

ggally_cor(diamonds, mapping = aes(x = depth, y = price))
```


```{r}

# price v carat boxplot. cut_width necessary as carat is also continuous. 
diamonds %>% 
  ggplot() +
  geom_boxplot(aes(x = carat, y = price, group = cut_width(carat, .25)))

# price v color (D is high, J is low)... appears to have an inverse relationship with price
diamonds %>% 
  ggplot() +
  geom_boxplot(aes(color, price)) 

# price v cut
diamonds %>% 
  ggplot() +
  geom_boxplot(aes(cut, price)) 

# price v clarity (I1 to IF)... appears to have an inverse relationship with price
diamonds %>% 
  ggplot() +
  geom_boxplot(aes(clarity, price)) 

diamonds %>% 
  ggplot() +
  geom_boxplot(aes(cut, carat)) 

#negative relationship between cut and carat... does this imply that larger carat diamonds are more likely to have inferior cuts. So price is driven more by carat which also happens to be correlated to inferior cut?

```

# 3. Install the ggstance package, and create a horizontal boxplot. How does this compare to using coord_flip()?

```{r}
# Brett Solution
diamonds %>%
  ggplot(aes(x = price, y = cut)) +
  geom_boxploth()

diamonds %>% 
  ggplot(aes(x = price, y = cut)) +
  geom_boxplot()

# Mike Solution
diamonds %>% 
  ggplot(aes(carat, cut, fill = color)) +
  geom_boxplot() 

diamonds %>% 
  ggplot(aes(carat, cut, fill = color)) +
  geom_boxploth()

# requires you to think about your x and y, and that they will not be flipped. 

```

# 4. One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using geom_lv() to display the distribution of price vs cut. What do you learn? How do you interpret the plots?

```{r}

diamonds %>% 
  ggplot(aes(x = cut, y = price)) +
  geom_lv()

ggplot (data = diamonds) +
  geom_boxplot(mapping = aes(x = cut, y = price)) + 
  geom_lv(aes(x = cut, y = price))

#demonstrates the distribution tapers as the price increases.

diamonds %>% 
  ggplot(aes(cut, price)) +
  geom_lv(aes(fill = ..LV..)) +
  coord_flip()
```

# 5. Compare and contrast geom_violin() with a facetted geom_histogram(), or a coloured geom_freqpoly(). What are the pros and cons of each method?

```{r}

diamonds %>% 
  ggplot(aes(carat, cut)) +
  geom_violin()

diamonds %>% 
  ggplot(aes(carat)) +
  geom_histogram() +
  coord_flip() +
  facet_wrap(~cut)

diamonds %>% 
  ggplot(aes(carat, color = cut)) +
  geom_freqpoly()

```

# 6. If you have a small dataset, it’s sometimes useful to use geom_jitter() to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to geom_jitter(). List them and briefly describe what each one does.

```{r}
# geom_quasirandom jitters your points in such a way that they are distributed meaningfully. Can be adjusted with method = frowney/smiley/tukey texture/tukey dense

#Same with geom_beeswarm.

ggplot(data = mpg) +
  geom_jitter(mapping = aes(
    x = reorder(class, hwy, FUN = median),
    y = hwy
  ))

ggplot(data = mpg) +
  geom_quasirandom(mapping = aes(
    x = reorder(class, hwy, FUN = median),
    y = hwy
  ))
```

-----

# 7.5.2 Two categorical variables

## Count between two categorical variables (instances that two categoricals appear together [ex F color, Ideal cut]) can be analyzed using dplyr::count or geom_count()

## geom_tile() can also be used if the count has been mapped using dplyr. 

# Exercises 7.5.2.1

# 1. How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut?


```{r}

#Mike Solution
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
  geom_tile(mapping = aes(fill = n)) +
  scale_fill_gradient2(low = "green",
                       mid = "blue",
                       high = "purple",
                       midpoint = 2000,
                       breaks=seq(0,4500,50), 
                       limits=c(1,5000))

# Steve Solution
diamonds %>%
  count(color, cut) %>%
  group_by(color) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(mapping = aes(x = color, y = cut)) +
  geom_tile(mapping = aes(fill = prop))

```

# 2. Use geom_tile() together with dplyr to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?

```{r}
#Mike Solution
flights %>% 
  group_by(dest, month) %>% 
  summarise(average_delay = mean(dep_delay, na.rm = T)) %>% 
  ggplot(aes(x = factor(month), y = dest)) +
  geom_tile(aes(fill = average_delay))

#Brett Solution
flights %>% 
  group_by(dest, month, dep_delay) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(dest, dep_delay, fill = month)) +
  geom_tile()

```

# 3. Why is it slightly better to use aes(x = color, y = cut) rather than aes(x = cut, y = color) in the example above?

```{r}

```


-----

# 7.5.3 Two Continuous Variables

Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width() vs cut_number()? How does that impact a visualisation of the 2d distribution of carat and price?


```{r}
ggplot(data = diamonds, mapping = aes(x = carat)) + 
  geom_freqpoly(mapping = aes(color = cut_number(carat, 20)))

# When using cut_number(), specify the # number of bins and the widths will be calculated automatically.

ggplot(data = diamonds, mapping = aes(x = carat)) + 
  geom_freqpoly(mapping = aes(color = cut_width(carat, 20)))

# When using cut_width(), choose the # width and the number of bins will be calculated automatically.

```

Visualise the distribution of carat, partitioned by price.


```{r}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price, color = cut_number(price, 10))) +
  geom_smooth(SE=FALSE)

ggplot(data = diamonds,
       mapping = aes(x = cut_number(price, 10), y = carat)) +
  geom_boxplot() +
  coord_flip()

diamonds %>% 
  ggplot(aes(x = carat, y = price, color = cut_number(price, 7))) +
  geom_boxplot()

diamonds %>% 
  ggplot() +
  geom_boxplot(aes(carat, price, group = cut_width(price, 1000)))

```

How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?

```{r}



```


Combine two of the techniques you’ve learned to visualise the combined distribution of cut, carat, and price.

Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of x and y values, which makes the points outliers even though their x and y values appear normal when examined separately.
