---
title: "Chapter 10 - Tibble"
author: "Mike McCann"
date: "9/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggplot2)
```

### Tibbles

## 10.1 Introduction

Tibbles are the updated version of data.frames and work better with modern packages. Usually these terms are used interchangeably. 

## 10.2 Creating Tibbles

- as_tibble() to coerce data.frame into a tibble. 
- tibble() create new tibbles from individual vectors.
- tribble() "transposed tibble".. entries separated by commas. 

## 10.3 Tibbles vs data.frame

Two main differences between tibble and data.frame: printing and subsetting

10.3.1 Printing. 
Tibbles (by default) only show the first 10 rows and only the columns that fit on screen. Keep from overloading the console.
-Use print(n = XX, width = Inf) to overide defaults

10.3.2
- $ extract by name 
    ex: df$x
- [[]] extract by name or position
    df[['x']] or df[[1]]

## 10.4 Interacting with older code

some older functions don't function with tibbles. Use as.data.frame() to convert back to data.frame.

## 10.5 Exercises

#1 How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).

```{r}

# Steve:
  
as_tibble(mtcars)

is_tibble(mtcars)

is_tibble(as_tibble(mtcars))

class(mtcars)

class(as_tibble(mtcars))

# ----

# Mike
  
print(mtcars)

as_tibble(mtcars)

# tibbles doesn't print all the columns when 'printed' (see initial row). 
# you can also use the function is_tibble()

is_tibble(mtcars)

```
 
#2 Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

```{r}
# Kayla
df <- data.frame(abc = 1, xyz = "a")
df$x
# [1] "a"
df[, "xyz"]
# [1] "a"
df[, c("abc", "xyz")]
#displays columns as <dbl>

df_tibble <- as_tibble(df)
df_tibble
df_tibble$x
#*** Unknown or uninitialised column: x
#changing to df_tibble$xyz returns the proper column. difference in interaction between $ and tibble v df.
df_tibble[, "xyz"]
#returns a without the quotes
# Returns rows from "xyz" column as <chr>
df_tibble[, c("abc", "xyz")]
#same ?

# ----

#Mike

df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

df2 <- as_tibble(df)
df2$x
df2[, "xyz"]
df2[, c("abc", "xyz")]

#tibble doesnt get past step two because it doesnt recognize x as == to xyz like it did with the dataframe.

```


#3 If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?

```{r}
#  Brett:
mtcars$mpg

tibble_mtcars <- as.tibble(mtcars)
var <- 'mpg'
tibble_mtcars[var]

# ----

#  Mike:

#using double brackets [[]] 

```

#4 Practice referring to non-syntactic names in the following data frame by:

Extracting the variable called 1.
Plotting a scatterplot of 1 vs 2.
Creating a new column called 3 which is 2 divided by 1.
Renaming the columns to one, two and three.


```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

## Major takeaway... ` should be used around numerical column names/variables to call them... 


# Kayln:
annoying$"1"

# Steve:
ggplot(annoying, aes(x = 1, y = 2)) +
  geom_point()

ggplot(annoying, aes(x = `1`, y = `2`)) +
  geom_point()

# Kayla
(annoying <- mutate(annoying, `3` = `2` / `1`))

# Brett
annoying %>% 
  rename("one" = `1`,
         "two" = `2`,
         "three" = `3`)

# ----

# Mike:

#Extracting the variable called 1.
annoying[["1"]]

#Plotting a scatterplot of 1 vs 2.

#does not work
annoying %>% 
  ggplot(aes(x = '1', y = '2')) +
  geom_point()

#does
annoying %>% 
  ggplot() +
  geom_point(aes(x = annoying$'1', y = annoying$'2'))

# Creating a new column called 3 which is 2 divided by 1.
annoying <- annoying %>% 
  mutate(`3` = `1` / `2`)

annoying %>% 
  rename(one = `1`, two = '2', three = '3')
```

#5 What does tibble::enframe() do? When might you use it?

```{r}

# enframe() converts named atomic vectors or lists to one- or two-column data frames.

enframe(1:3)
enframe(c(a = 5, b = 7))

```

#6 What option controls how many additional column names are printed at the footer of a tibble?

```{r}
# # Steve:
# n_extra
# Number of extra columns to print abbreviated information for, if the width is too small for the entire tibble. If NULL, the default, will print information about at most tibble.max_extra_cols extra columns.

# ----

# Mike:

mttibble <- as_tibble(mtcars)

print(mttibble)

#  vs

print(mttibble, n_extra = 3)
```

